---
title: "260219 - Generating GO pathway heatmaps"
output:
  html_document:
    df_print: paged
---
***Purpose:***
The purpose of this script is to create a heatmap that displays the z-score standardized gene expression of each differentially regulated gene of a selected pathway across all samples (separated between groups). You will need two spreadsheets:
- the pathway enrichment sheet (e.g. GO bp)
- the DEG list

***Important steps:***
1. You need to place the same metadata you have been using for previous scripts into the directory
2. Place the normalized counts from the very first normalization script into the "Normalized_Counts" folder inside the "Result_sheets" folder (just as you did for previous scripts)
3. Place the same combined GOterms files that you used for the previous top 10 dotplot script into the "GOterms" folder inside the "Result_sheets" folder 
4. Update the following (this isn't an automated script because it is all about displaying what you actually want):
- the organism database you need (org.Hs.eg.db or org.Mm.eg.db)
- the file names for your spreadsheets
- the name of the pathway that you want
- the ontology (BP, CC, or MF)
- the group names (add and subtract as needed in two parts of the script)
- the save file name

***You will need to update the following before running again:***
- the file names for your spreadsheets
- the name of the pathway that you want
- the sample names 
- where you want the plots saved

***Output (what you will get):***
- You will get a single pdf file with a single heatmap for the pathway that you choose

***What you can adjust:***
- Can change the color scheme

*NOTE*
Have a look at how many genes are in the pathway you want to look at, often they can have around 100 genes which makes a very large plot and you may need to adjust the file save size to make it fit in a single pdf. I try to use more specific pathways that have fewer genes for a more manageable plot.

Here is the full script:
```{r}
# Load necessary libraries
library(ggplot2)
library(pheatmap)
library(dplyr)
library(tidyr)
library(tibble)
library(clusterProfiler)
library(org.Hs.eg.db) # Change for human or mouse
library(readxl)

###

#Reading in the CSV files that I saved into the directory (Need to change this based on what files you are using)
pathway_results <- read_excel("Result_sheets/GOterms/Membrane_vs_Blood_deg_GO_enrichment_results_combined.xlsx")  # Replace with your file path

gene_expression_data <- read.csv("Result_sheets/Normalized_Counts/normalized_counts.csv",
                            header = TRUE,
                            row.names = 1)  # Replace with your file path

metadata <- read.csv("metadata.csv") # Replace with your file path

# Change the rownames of the metadata column to the sample names
rownames(metadata) <- paste(metadata$Group, metadata$Replicate, 
                         sep = "_")

# Change the column names of the count data to the sample names
colnames(gene_expression_data) <- rownames(metadata)

###

# Extract Ensembl IDs from row names
Ensembl.id <- rownames(gene_expression_data)

# Convert Ensembl IDs to gene names
gene_names <- bitr(Ensembl.id, fromType = "ENSEMBL", toType = "SYMBOL", OrgDb = org.Hs.eg.db) # Change for human or mouse

# Merge the gene names with your original data based on Ensembl IDs
gene_expression_data$Symbol <- gene_names$SYMBOL[match(rownames(gene_expression_data), gene_names$ENSEMBL)]
# Select a pathway of interest (adjust the pathway name as needed)
selected_pathway <- "positive regulation of angiogenesis"  # Replace with the name of your pathway of interest

###

# Filter the GO pathway data to get the list of genes for the selected pathway
genes_in_pathway <- pathway_results %>%
  filter(Description == selected_pathway, GeneSet == "Upregulated_genes", Ontology == "BP") %>% #Can change Downregulated to Upregulated or All
  pull(geneID) %>%
  strsplit(split = "/") %>%
  unlist()

# Filter the gene expression data to keep only the genes in the selected pathway
gene_expression_filtered <- gene_expression_data %>%
  filter(Symbol %in% genes_in_pathway)

# Sort genes alphabetically
gene_expression_filtered <- gene_expression_filtered %>%
  arrange(Symbol)

# Select the control and experimental sample columns and order them accordingly
sample_columns <- gene_expression_filtered %>%
  dplyr::select(starts_with("Blood"), starts_with("Membrane")) %>%
  colnames()

# Extract the count data for these samples
count_data <- gene_expression_filtered %>%
  dplyr::select(Symbol, all_of(sample_columns))

###

# Set gene names as rownames and remove the Symbol column from data
heatmap_data <- count_data
rownames(heatmap_data) <- heatmap_data$Symbol
heatmap_data <- heatmap_data[, -1]  # Remove the Symbol column

# Convert to matrix
heatmap_matrix <- as.matrix(heatmap_data)

# Calculate z-scores for each row (gene)
z_score_matrix <- t(apply(heatmap_matrix, 1, function(x) (x - mean(x)) / sd(x)))

###

# Clean up the pathway name for the file (remove spaces, special characters)
cleaned_pathway_name <- gsub("[^A-Za-z0-9]", "_", selected_pathway)

# Define the sizes of each group (change based on the samples you used)
group_sizes <- c(
  sum(startsWith(sample_columns, "Membrane")), 
  sum(startsWith(sample_columns, "Blood"))
)

# Set up the PDF device for saving (change the comparison and the direction each time)
pdf(file = paste0("Plots/", cleaned_pathway_name, "_MembranevsBlood", "_Up", "_heatmap.pdf"), width = 10, height = 12)

# Plot heatmap using pheatmap, with custom gaps and thicker borders
pheatmap(z_score_matrix, 
         cluster_rows = TRUE,  # Cluster rows (genes)
         cluster_cols = FALSE,  # Do not cluster columns (samples)
         show_rownames = TRUE,  # Show gene names on rows
         show_colnames = TRUE,  # Show sample names on columns
         main = paste("Heatmap of Z-Scores for Pathway:", selected_pathway),
         gaps_col = cumsum(group_sizes)[-length(group_sizes)],  # Adds a visual gap between groups (Assigned above)
         border_color = "black",  # Black border around all cells
         cellwidth = 15,  # Adjust cell width to make the gap more visible
         cellheight = 12,  # Increase cell height to fit gene names
         fontsize = 8,  # Reduce font size for overall plot
         fontsize_row = 10,  # Smaller font size for row names (gene names)
         fontsize_col = 10,  # Font size for column names (sample names)
         angle_col = 45,  # Rotate column names to prevent overlap
         color = colorRampPalette(c("#4682B4", "white", "#E34234"))(50),  # Custom color palette
         legend_breaks = seq(-3, 3, 1),  # Define the scale for the legend
         legend_labels = seq(-3, 3, 1),  # Label the scale
         )

# Turn off the plotting device
dev.off()
  
```


********************************************************************************


I like to finish with the session info command as it lists all the versions that were used at the time of analysis. The other lines will save the output as a text file with the date that the command was run.
```{r}
sessionInfo()

# Create filename with date prefix
filename <- paste0(Sys.Date(), "_sessionInfo.txt")

# Save sessionInfo output
capture.output(sessionInfo(), file = filename)

```
