---
title: "260212 - KEGG Enrichment Analysis_Novogene"
output: html_notebook
---
***Purpose:***
This script is designed to run KEGG enrichment analysis using the DEG spreadsheets that you get from the Novogene data package. *These are the same that you use for the volcano plots.* 

This uses the clusterProfiler package which is the gold standard for robust and repeatable pathway analysis at the momement.

#Note that KEGG requires the use of Entrez ID and has a smaller database of genes, so don't be alarmed when less genes make it through to the analysis

***Important steps:***
1. You need to take the DEG spreadsheets from Novogene that you used in the volcano plot script and place them in the "DEG_Spreadsheets" folder.
2. Make sure that the right organism database is being used. (e.g. org.Hs.eg.db for humans and org.Mm.eg.db for mice). 
3. Run the script, it should automatically do the analysis for you. Note that the full automated script can take a while (minutes) depending on how many spreadsheets you are looping through.

***Output (what you will get):***
- For each comparison, you will get a full .csv file that has all of the significant KEGG pathways from every combination of up/down/all DEGs (This is used in a future script)
- You will also get a folder for each comparison that has these different combinations in separate spreadsheets. This is better for actually looking through the data.

***What you can adjust:***
- You can adjust the logFC cutoff to change the amount of genes you input into the pathway analysis. Be sure to change the output folder and file names to reflect this. You should publish volcano plots and pathway analysis with a consistent threshold so if you change this, you should rerun your volcano plots with the appropriate threshold lines.
- You can play around with various settings in the analysis, e.g. statistical methods

Here is the full script:
```{r}
# Load necessary libraries
library(clusterProfiler)
library(org.Hs.eg.db) # Change for mouse or human
library(openxlsx)
library(dplyr)

# Define input and output directories
input_dir <- "DEG_Spreadsheets"
output_dir <- "Results_KEGG"

dir.create(output_dir, showWarnings = FALSE)

# List all CSV files
files <- list.files(input_dir, pattern = "*.csv", full.names = TRUE)


###

# KEGG enrichment function
run_kegg <- function(gene_list, gene_set_name, subfolder, file_prefix) {

  # Skip tiny gene sets
  if (length(gene_list) < 5) return(NULL)

  kegg_res <- tryCatch(
    enrichKEGG(
      gene = gene_list,
      universe = all_genes,
      organism = "hsa", # mmu for mouse or hsa for human
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.05
    ),
    error = function(e) NULL
  )

  if (is.null(kegg_res)) return(NULL)

  kegg_res <- setReadable(kegg_res, OrgDb = org.Hs.eg.db, keyType = "ENTREZID") # KEGG uses entrez ID, so this step switches to gene symbol also change database for mouse or human
  result_df <- as.data.frame(kegg_res)

  # 🚨 No significant pathways
  if (nrow(result_df) == 0) return(NULL)

  # Add metadata
  result_df$GeneSet <- gene_set_name
  result_df$Database <- "KEGG"

  # Save individual file
  individual_file <- file.path(
    subfolder,
    paste0(file_prefix, "_", gene_set_name, "_KEGG.xlsx")
  )

  write.xlsx(result_df, individual_file, rowNames = FALSE)

  return(result_df)
}


###

# Main loop through all files (combinations)
for (file in files) {

  master <- read.csv(file)
  file_prefix <- tools::file_path_sans_ext(basename(file))

  # Convert all genes to Entrez
  all_genes <- bitr(
    master$gene_id,
    fromType = "ENSEMBL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db # Change for mouse or human
  ) %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unique()

  All.degs <- bitr(
    master$gene_id[master$padj < 0.05],
    "ENSEMBL", "ENTREZID", org.Hs.eg.db # Change for mouse or human
  ) %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unique()

  upregulated <- bitr(
    master$gene_id[master$padj < 0.05 & master$log2FoldChange > 1],
    "ENSEMBL", "ENTREZID", org.Hs.eg.db # Change for mouse or human
  ) %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unique()

  downregulated <- bitr(
    master$gene_id[master$padj < 0.05 & master$log2FoldChange < -1],
    "ENSEMBL", "ENTREZID", org.Hs.eg.db # Change for mouse or human
  ) %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unique()

  # Output folder per comparison
  subfolder <- file.path(output_dir, file_prefix)
  dir.create(subfolder, showWarnings = FALSE)

  all_results <- list()

  all_res  <- run_kegg(All.degs, "All_genes", subfolder, file_prefix)
  if (!is.null(all_res)) all_results <- append(all_results, list(all_res))

  up_res   <- run_kegg(upregulated, "Upregulated_genes", subfolder, file_prefix)
  if (!is.null(up_res)) all_results <- append(all_results, list(up_res))

  down_res <- run_kegg(downregulated, "Downregulated_genes", subfolder, file_prefix)
  if (!is.null(down_res)) all_results <- append(all_results, list(down_res))

  # Save combined KEGG results
  if (length(all_results) > 0) {
    final_results <- do.call(rbind, all_results)
    output_file <- file.path(
      output_dir,
      paste0(file_prefix, "_KEGG_enrichment_results_combined.xlsx")
    )
    write.xlsx(final_results, output_file, rowNames = FALSE)
  }
}
```


********************************************************************************


I like to finish with the session info command as it lists all the versions that were used at the time of analysis. The other lines will save the output as a text file with the date that the command was run.
```{r}
sessionInfo()

# Create filename with date prefix
filename <- paste0(Sys.Date(), "_sessionInfo.txt")

# Save sessionInfo output
capture.output(sessionInfo(), file = filename)

```